Species motif participation provides unique information about species risk of extinction

### 2023-04-18 manuscript/reorder_main_text.tex

    A species' position within a food web (a network of who eats whom) is known to affect its ability to persist following the extinction of  another species in the network (i.e., secondary extinction)~\citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    For example, species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally more likely to go extinct~\citep{binzer2011susceptibility, Eklof2006}.
    Species with low in-degree are more vulnerable to extinction because they have few alternative resources if a prey species is lost, while those with high trophic levels are vulnerable to the loss of any species along the path from basal resource(s) to the focal species.
    Based on these trends, it has also been proposed that a species' participation in \emph{meso-scale structures} (which take into account the species' direct and indirect interactions) are also related to its persistence.

    One way to describe a species' participation in such meso-scale structures is using its participation in the set of \emph{motifs} of a given size~\citep{Cirtwill2015a,Cirtwill2018FoodWebs}.
    Motifs --unique patterns of $n$ interacting species-- correspond to different sets of direct and indirect interactions and provide a holistic description of a species' place in a network~\citep{Stouffer2007,Stouffer2012}.
    Like degree and trophic level, a species' \emph{motif participation} (the vector of frequencies with which it appears in each motif) can easily be calculated once the structure of the network is known.
    Unlike in-degree and trophic level, which focus only on direct prey of the focal species or on the lengths of paths between the focal species and basal resources, motif participation also incorporates information on the focal species' interaction partners.
    Two species might both have three prey and a trophic level of 3, but they might have quite different motif participation if one species's prey are strict specialists and the others' prey are generalists.
    Importantly, since specialists are generally more vulnerable to extinction than generalists, the predator of specialists will be more likely to lose prey and go extinct itself than a predator of generalists.

    While a relationship between motif participation and persistence is broadly intuitive, so far there has been little research demonstrating which motifs are associated with greater probabilities of persistence.
    Here we focus on persistence after the loss of basal resources (i.e., bottom-up disturbances).
    The plant community is the foundation upon which the myriad species in food webs depend for their survival.Disturbances to the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
    The loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}.
    Species' motif participation may indicate \emph{which} consumers are more or less likely to go extinct following such a disturbance.

    One potential reason for the difficulty in assessing the relationship between species' motif roles and persistence are the large computational resources required to simulate population dynamics.
    Our solution is to use Bayesian networks, which are much more computationally efficient and facilitate the analysis of larger networks~\citep{Eklof2013,Haussler2020}.
    Although Bayesian networks currently only simulate bottom-up disturbances (our focal interest), they nevertheless capture the vast majority of secondary extinctions obtained from dynamical models including top-down and bottom-up effects~\citep{Eklof2013}.
    The greater computational efficiency and similar [[results? performance? extinctions?]] of Bayesian networks allow us to simulate the effects of a gradient of disturbance strength and thereby explore whether participation in some motifs is associated with greater probability of extinction under strong or weak disturbances.

    Although there are 13 possible three-species motifs (the most commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}, we focus on just four: the three-species chain, apparent competition, direct competition, and omnivory.
    These four motifs are by far the most common in empirical networks~\citep{Stouffer2007, Borrelli2015a, giling2019plant}.
    They are also the only motifs which do not include cycles (where species A and B eat each other or where A eats B eats C eats A).
    As modelling bottom-up extinctions in a Bayesian framework requires the breaking of cycles, these are the only motifs that will appear in the resulting \emph{acyclic} networks.
    Fortunately, \cite{Allesina2009functional} proved that cyclic and acyclic versions of a network have the same properties in terms of secondary extinctions and the effect on network robustness are very limited.

    Moreover, these four motifs have clear ecological interpretations, which are lacking for the cyclical motifs.
    The three-species chain indicates vertical flows of energy and biomass in a network and captures the possibility of indirect interactions across multiple trophic levels (e.g., between a plant and a predator of the plant's herbivore).
    The direct competition motif, where two consumers share the same resource, captures the possibility of indirect interactions between consumers (e.g., by depletion of the shared resource). 
    The apparent competition motif, where a predator consumes two resources, can include indirect effects between prey (e.g., by boosting predator populations) in a top-down context, but in a bottom-up context only includes direct interactions.
    Finally, the omnivory motif, where a predator consumes the same resource as its prey, includes captures the potential for a plant to have both direct and indirect effects on the predator.

@@

@@

        Species' motif participation varied depending on the global properties of the network: size, connectance, and their interaction (Fig. S1, Table S2; \emph{Appendix S5}).
        Species appeared most often in the apparent competition motif in all types of networks, but the proportion of apparent competition decreased with increasing network size and connectance.
        The proportion of the omnivory motif in a species' participation role increased sharply with increasing connectance, while the proportion of the direct competition motif increased slightly with increasing network size.

        Species' motif participation also varied with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
        Species with higher in-degree tended to have higher frequencies of participation in the omnivory motif and lower proportions of the other three motifs.
        Species with higher trophic levels tended to have higher proportions of apparent competition and three-species chains and lower proportions of omnivory and direct competition. 
        A species' persistence varies with both species in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}; \emph{Appendix S6}). 
        Increased trophic level is generally associated with decreasing species persistence, although the rate of decline is slower at high disturbance (Table S3, \emph{Appendix S6}).
        In-degree shows an even stronger interaction with disturbance; having more prey species is associated with higher persistence without disturbance, but lower persistence at high levels of disturbance.

        \subsubsection*{Consistency in effects of motif participation on species persistence}

            When fitting regressions of persistence against motif participation for each network separately, networks with high connectance generally showed more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}).
            However, this trend varied at different levels of disturbance, especially for the direct competition and omnivory motifs.
            At low and medium connectances there was little change in the distribution of slopes for direct competition across all levels of disturbance. 
            At high connectances, however, the proportion of positive slopes was both lower overall and decreased with increasing disturbance. 
            For the omnivory motif these trends were reversed. 
            Moreover, the omnivory motif has a particularly broad distribution of slopes (i.e., an inconsistent relationship to persistence) when network connectance was low. 
            Network size had a much smaller effect than connectance (Fig. S4; \emph{Appendix S7}).

    \subsection*{Network mean persistence and motif profiles}

        The overall motif profile of a network was related to the average persistence of consumers ($F_{1,2998}$=645, $p$\textless0.001), although this result must be considered tentative due to non-homogeneous distribution of roles (\emph{Appendix S8}). 
        Average persistence was higher in networks with higher proportions of the apparent competition motif and lower in networks with higher proportions of the omnivory motif (Table S4, \emph{Appendix S8}). 
        These trends were consistent across levels of disturbance (Fig.~\ref{fig:motif_profile_persistence}, Table S4), \emph{Appendix S8}).

    \subsection*{Motif profiles and network properties}

        Considering each motif separately, the proportion of the omnivory motif in a network's motif profile increased significantly with increasing connectance (Fig.~\ref{motif_proportion_lms}, Table S5, \emph{Appendix S9}).
        The proportions of other motifs in a network's profile were not related to network size, connectance, or their interaction.
        Consistent with the results of these GLMMs, a PERMANOVA test suggested that a network's motif profile as a whole was related to network size and connectance.
        However, the assumption of equal variance across groups was violated for this test and the PERMANOVA results should therefore be considered indicative only.

>>>>

    The loss of a species in a food web can, due to the mutual dependencies among species,  trigger a cascade of additional extinctions, so called secondary extinctions. A species' position within a food web is known to affect its risk of going secondarily extinct ~\citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    For example, species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally more likely to go extinct~\citep{binzer2011susceptibility, Eklof2006}.
    Species with low in-degree are more vulnerable to extinction because they have few alternative resources if a prey species is lost, while those with high trophic levels are vulnerable to the loss of any species along the path from basal resource(s) to the focal species.
    Based on these trends, it has also been proposed that a species' participation in \emph{meso-scale structures} (which take into account the species' direct and indirect interactions) are also related to its persistence.

    One way to describe a species' participation in such meso-scale structures is using its participation in the set of \emph{motifs} of a given size~\citep{Cirtwill2015a,Cirtwill2018FoodWebs}.
    Motifs --unique patterns of $n$ interacting species-- correspond to different sets of direct and indirect interactions and provide a holistic description of a species' place in a network~\citep{Stouffer2007,Stouffer2012}.
    Like degree and trophic level, a species' \emph{motif participation} (the vector of frequencies with which it appears in each motif) can easily be calculated once the structure of the network is known.
    However, unlike in-degree and trophic level, which focus only on direct prey of the focal species or on the lengths of paths between the focal species and basal resources, motif participation also incorporates information on the focal species' interaction partners.
    Two species might both have three prey and a trophic level of 3, but they might have quite different motif participation if one species's prey are strict specialists and the others' prey are generalists.
    Importantly, since specialists are generally more vulnerable to extinction than generalists, the predator of specialists will be more likely to lose prey and go extinct itself than a predator of generalists.

    The plant community is the foundation upon which the myriad species in food webs depend for their survival. Disturbances to the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
    The loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}. Given the severity of disturbances, here loss of species, at the lowest trophic levels, we here choose to focus on persistence after the loss of basal resources (i.e., effects of bottom-up disturbances).
    Species' motif participation may indicate \emph{which} consumers are more or less likely to go extinct following loss of primary producers.

    While a relationship between motif participation and persistence is broadly intuitive, so far there has been little research demonstrating which motifs are associated with greater probabilities of persistence.
    One potential reason for the difficulty in assessing the relationship between species' motif roles and persistence are the large computational resources required to simulate population dynamics.
    Our solution is to use Bayesian networks, which are much more computationally efficient and facilitate the analysis of larger networks~\citep{Eklof2013,Haussler2020}.
    One limitation with Bayesian networks is that they only take into account bottom-up effects (see Methods). Nevertheless, Bayesian networks do capture the vast majority of secondary extinctions obtained from fully-fledged dynamical models including top-down and bottom-up effects~\citep{Eklof2013}, and as such are a good proxy for analysing species persistence following extinctions of primary producers in complex food webs.
    %The greater computational efficiency and similar [[results? performance? extinctions?]] of Bayesian networks allow us to simulate the effects of a gradient of disturbance strength and thereby explore whether participation in some motifs is associated with greater probability of extinction under strong or weak disturbances.

    %There are 13 possible three-species motifs (the most commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}, 
    We will here focus three-species motifs, which are the commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}. We particularly analyze four of the three-species motifs, namenly the three-species chain, apparent competition, direct competition, and omnivory.
    We choose these four motifs as they are by far the most common in empirical networks~\citep{Stouffer2007, Borrelli2015a, giling2019plant}.
    These four motifs also have clear ecological interpretations:
    The three-species chain indicates vertical flows of energy and biomass in a network and captures the possibility of indirect interactions across multiple trophic levels (e.g., between a plant and a predator of the plant's herbivore).
    The direct competition motif, where two consumers share the same resource, captures the possibility of indirect interactions between consumers (e.g., by depletion of the shared resource). 
    The apparent competition motif, where a predator consumes two resources, can include indirect effects between prey (e.g., by boosting predator populations) in a top-down context, but in a bottom-up context only includes direct interactions.
    Finally, the omnivory motif, where a predator consumes the same resource as its prey, includes captures the potential for a plant to have both direct and indirect effects on the predator. 
    Furthermore, these four motifs do not include cycles (where species A and B eat each other or where A eats B eats C eats A) and as modelling secondary extinctions in a Bayesian network requires acyclic networks (see Methods), these are the only motifs that will appear in the resulting \emph{acyclic} networks.

     %They are also the only motifs which do not include cycles (where species A and B eat each other or where A eats B eats C eats A).
    %As modelling bottom-up extinctions in a Bayesian framework requires the breaking of cycles, these are the only motifs that will appear in the resulting \emph{acyclic} networks.
    %Fortunately, \cite{Allesina2009functional} proved that cyclic and acyclic versions of a network have the same properties in terms of secondary extinctions and the effect on network robustness are very limited.

@@

    \subsection*{Species persistence and other network properties}

        % The overall motif profile of a network was related to the average persistence of consumers ($F_{1,2998}$=645, $p$\textless0.001), although this result must be considered tentative due to non-homogeneous distribution of roles (\emph{Appendix S8}).
        As well as the frequency with which each species appears in each motif, we can use the total frequencies of each motif in the whole network (i.e., the network's motif profile) to describe its structure.
        The average probability of persistence of all consumers in the network varied with the frequencies of different motifs in this profile.
        Specifically, average consumer persistence was higher in networks with higher proportions of the apparent competition motif, lower in networks with higher proportions of the omnivory motif, and was not significantly related to the frequencies of the three-species chain or direct competition motifs (Fig.~\ref{fig:motif_profile_persistence}, Table S4, \emph{Appendix S8}). 
        There was no significant interaction between the proportion of any motif in the network profile and disturbance (Table S4), \emph{Appendix S8}).

        To provide context for the relationships between consumer persistence and motif participation or motif profile, we also tested whether persistence was related to more-familiar network properties.
        Average consumer persistence was also lower in networks with higher connectance but was not significantly related to network size (Table S6, \emph{Appendix S10}).
        As expected, a consumer's persistence tended to decrease with increasing STL (Fig.~\ref{fig:motifs_vs_TL_and_deg}, Table S3, \emph{Appendix S6}).
        The relationship between in-degree and persistence depended on the strength of disturbance: higher in-degree was associated with greater persistence when basal species had low probability of extinction but lower persistence when basal species were highly likely to go extinct.

@@

        Both the motif profile of a network and species' motif participation were related to other measures of network structure.
        Most simply, the proportion of the omnivory motif in the network's motif role increased with increasing connectance (Fig.~\ref{motif_proportion_lms}; Table S5, \emph{Appendix S9}).
        The proportions of other motifs did not vary significantly with connectance and none were significantly related to network size.

        A species' motif participation varied with network size, connectance, and their interaction. 
        The proportion of the omnivory motif generally increased with increasing connectance and network size, but decreased with increasing network size in the most-connected networks (Fig. S2 and Table S2, \emph{Appendix S5}).
        The proportion of apparent competition decreased strongly with increasing connectance and more weakly (but still significantly) with increasing network size.
        The proportion of the three-species chain showed similar trends to apparent competition, while the proportion of direct competition showed very little change over the range of network size and connectance we consider (though the effects of connectance and the interaction between connectance and network size were significant).

        Species' motif participation also varied with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
        Species with higher in-degree tended to have higher frequencies of participation in the omnivory motif and lower proportions of the other three motifs.
        Species with higher trophic levels tended to have higher proportions of apparent competition and three-species chains and lower proportions of omnivory and direct competition. 
        A species' persistence varies with both species in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}; \emph{Appendix S6}).

        % \subsubsection*{Consistency in effects of motif participation on species persistence}

        %     When fitting regressions of persistence against motif participation for each network separately, networks with high connectance generally showed more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}).
        %     However, this trend varied at different levels of disturbance, especially for the direct competition and omnivory motifs.
        %     At low and medium connectances there was little change in the distribution of slopes for direct competition across all levels of disturbance. 
        %     At high connectances, however, the proportion of positive slopes was both lower overall and decreased with increasing disturbance. 
        %     For the omnivory motif these trends were reversed. 
        %     Moreover, the omnivory motif has a particularly broad distribution of slopes (i.e., an inconsistent relationship to persistence) when network connectance was low. 
        %     Network size had a much smaller effect than connectance (Fig. S4; \emph{Appendix S7}).
